---
title: "Predicting Fire Behavior Using Weather & Meteorological Data"
author: "Darshil Desai, Edwin Ramirez"
date: "November 27, 2018"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

# 2.1 Abstract

# 2.2 Introduction
Our regressional model will focus on utilizing previous forest fire and weather data sourced from the UCI Machine Learning Repository to predict the Initial Spread Index (ISI) of fires in Portugal and any countries using the Canadian Forest Wildfire Index System (CFWIS). We found this project to be of relevance considering the recent devastating droughts and fires in Northern Europe throughout Greece, Portugal, and Spain within the last year. Additionally, the relevance of the recent California fires also give support to the necessity of predictive analytics in this area. Hence, by utilizing a regressional model, we can predict the early behavior of a fire. Portugal utilizies the Canadian Forest Fire Weather Index System in order to track fuel moisture and wind speed to determine the intensity of a fire. Thus, by predicting the initial spread index of potential fires, we'd be able to predict the danger of a fire based on how quickly it would spread. The ISI scale begins at 0, where a 10 indicates a high rate of spread after ignition, and 16 or higher indicates an extreme rapid rate of spread.   

The research question can be hypothesized as follows:

$H_{0}:$ None of the predictor variables in the dataset are useful in making predictions about the Initial Spread Index.

$H_{a}$: At least one of predictor variables in the dataset is useful in making predictions about the Initial Spread Index.


# 2.3 Dataset 

## View Dataset

The UCI dataset comprises of the following 13 variables: 

1 & 2. X,Y : coordinates within the Montesinho park. Ranges from 1 to 9 <br>
3. month: month when the fire frst occured <br>
4: day: day of the given month when the fire occured <br>
5: FFMC (Fine Fuel Moisture Code) : a numeric rating of the moisture content of litter and other cured fine fuels. This code is an indicator of the relative ease of ignition and the flammability of fine fuel. <br>
6: DMC (Duff Moisture Code): A numeric rating of the average moisture content of loosely compacted organic layers of moderate depth. This code gives an indication of fuel consumption in moderate duff layers and medium-size woody material.[^1] <br>
7.Burned area of the forest: area of forest burned  <br>
8. temperature <br>
9: RH (relative humidity) <br>
10: wind: wind speed <br>
11: rain <br>
12: DC (Drought Code):  A numeric rating of the average moisture content of deep, compact organic layers. This code is a useful indicator of seasonal drought effects on forest fuels and the amount of smoldering in deep duff layers and large logs. <br>
13. ISI (Initial Spread Index): expected rate of fire spread.This variable will be our response variable and we will try and establish a linear relationship between the myriad of weather and meterological factors of the forest experiencing fires and the future expected area burn. [^2] <br> 


``` {r see_data, warning = FALSE, echo = FALSE}
#reading in all the packages reqired for the report
library(ggcorrplot)
library(dplyr)
library(corrplot)
library(leaps)
library(dummies)
library(car)
library("psych")

#reading in the csv file
fire <- read.csv('forestfires.csv') 
head(fire)
```

## 2.4 Methodology
In order to draw a relationship between the various weather, meteorological variables in the dataset and the response variable (Initial Spread Index), it is vital to transform and preprocess our data to succesfully represent this data in a linear relationship. 

Our methods to do the same involve the following: <br>
1. Deleting redundant data that will fail to enhance the model's predictive power <br>
2. Representing categorical data (months) as binary numerical data. <br>
3. Transforming our response variable <br>
4. Examine multicolinearity <br>

### 1. Deleting Redundant Data
Several variables in the given dataset do not assist in creating the linear relationship between with the response variable (Initial Spread Index). Briefly:<br> 
a. X,Y: Coordinate data provided by the data can be considered redundant due to the coordintes being confined to the specific area of Montesinho Park in Portugal. This data can be removed because the goal of our model is to predict fires within countires that utilize the Canadian Fire Weather Index System.<br>
b. Days: In isolation, the numerical value of a day does not provide any significant relevance to a datapoint <br>

``` {r delete_data}
fire <- fire[,-match(c("day", 'X', 'Y'),names(fire))]
```


### 2. Categorical Data

The variable of months can be grouped to create a categorical variable that signifies seasons. Thus, each season will be represented as nominal variables in our regressional model. 
```{r data_cleaning_two}
spring <- c('mar', 'apr', 'may')
summer <- c('jun', 'jul', 'aug')
autumn <- c('sep', 'oct','nov')
winter <- c('dec','jan', 'feb')

#clean data
# Representing months as the appropriate season it is a part of
season = function(x){
  if(x %in% spring){
    return('spring')
  } else if(x %in% summer){
    return('summer')
  } else if(x %in% autumn){
    return('autumn')
  } else{
    return('winter')
  }
}

fire$month = sapply(fire$month, season)

#renaming the month column to "seasons"
colnames(fire)[colnames(fire)=="month"] <- "season_"

#Converting the months column into 4 columns, one for each season
fire <- dummy.data.frame(fire)
invisible(get.dummy( fire, 'season_'))

```

### 3.Transforming the response variable

The histogram below illustrates the distribution of the response variable as right-skewed. To prevent a normality violation in the regressional model, the data is transformed by taking the natural log of all ISI data.  

```{r show}
#check normality of response variable
hist(fire$ISI, col = "lightblue", xlab = "ISI", main = "Histogram of ISI")
```


```{r transform}
#transform data by taking log
trans = log(fire$ISI)

#change  the below case or error occurs
trans[380] = 1.2

hist(trans, xlab = "Transformed ISI", col = "lightblue", main = "Histogram of ISI Transformed")
abline(v = mean(trans), col = 'red')
abline(v = median(trans), col = 'blue')

```

Consequentially, the distribution the response variable (ISI) is now normal, and we can proceed with selecting the appropriate features for our model. A QQ-normal plot is shown below to illustrate that the residual data is normally distributed.  

```{r}
#use QQ plot check model
summary(lm(trans ~.-ISI, data = fire))
plot(lm(trans ~.-ISI, data = fire),2)
```

### 4.Examining Multicolinearity {.tabset}
Multicolinearity occurs when predictor variables are linearly correlated with the other. This implies that a change in any one of the predictor variables would entail a change in another highly correlated predictor variable. 

Before we proceed to fit our model it is important to perform two vital checks. Firstly, there needs to be a reasonable correlation between the predictor variables and the response variable. An absence of any linear pattern does not warrant the use of a linear regression problem.

#### Colinearity Between Predictor Variables

``` {r}
#Checking for multicolinearity. Here we will compare each of our predictor variables with all the others predictor variables. 
corr<-cor(fire)
#plotting correlation head map
corrplot(corr, type = 'upper')
```
#### Correlation Between Predictor and Response Variable 
``` {r analysis, warning = FALSE}
#1. Visualizing correlations between our response variable and the predictor variable 
# Comparing Y with all other predictor variables *(X's)
corr1<-cor(x = fire[,-8], y = trans, use="complete.obs")
ggcorrplot(corr1)
```

The visual analysis above shows us that there are several predictor variables that possess a significant correlation strength with the response variable (transformed ISI). Furthermore we also see that there exists no significant multicolinearity between the predictor variables. In general there always lies the possibility that variables within the same domain will correlate with one another to an extent. 

However to further confirm our claim, we will employ the use of the Variance Inflation Factor analysis. This analysis allows us to support / reject our claim using numerical proof. It is important to note that the lower the VIF (lowest being 1), the less multicolinearity exists in our dataset. A VIF of 5 represents industry standard acceptance rate for multicolinearity as a small value indicates that the standard deviation of the respective variable parameter will remain relatively stable when other predictor variables are added into the regression equation. 

``` {r vif}
#Calculate the Variance Inflation Factor for each of the predictor variables
model <- lm(trans~ fire$FFMC + fire$DMC + fire$DC + fire$temp + fire$RH + fire$wind + fire$rain + fire$area, data = fire)
#summary(model)
vif(model)

```
Based on the data above, it further supports that there exists no significant multicolinearity in the dataset. Most of the VIF values are under 2.5 far below the industry threshold of 5. 

# 2.5 Results & Discussion

After analyzing the correlation of the predictor variables in the dataset, the next steps comprise the following:  <br>
1. Feature Selection <br>
2. Model fitting using Training Data <br>
3. Analysis <br>

### 1.Feature Selection 

Using stepwise regression, we iterate through different possibilities of a linear model and choose one with features yileding the best (lowest) error metric.

The QQ-normal plot below illustrates the distribution of our data after utilizing stepwise regression for feature selection. 
```{r}
# forward stepwise regression
model = lm(trans ~.-ISI, data = fire)
new_model = step(model)

#new model QQ plot
plot(new_model, 2)

new_model$terms
```


### 2. Fitting the model

A portion of the data is separated to be utilized as a training set, while the remaining portion will be utilized as a test set.This will allow the accuracy of the model to be measured. 
``` {r}
#attaching the trans column to the main fire dataset
fire_to_split <- cbind(fire, trans)

#Get 400 random numbers
ran <- sample(1:517, 400, replace=F)

#splitting the dataset into training and testing
train <- fire_to_split[ran,]

#test data: contains data not in training data
test = setdiff(fire_to_split, train)
```


Next, the training data is fitted to the model with the selected features. After removing features not selected from stepwise regression, the model includes `season_autumn`,`season_spring` ,`season_summer`,`FFMC`, `temp`, `RH`, and `wind`
``` {r}
#fitting the model
m = lm(train$trans ~ season_autumn + season_spring + season_summer + FFMC + temp + RH + wind - ISI - trans, data = train)
summary(m)
```

### 3. Analysis

The following conclusions are the results of our model: 
- The overall P-value of the model is <2.2e-16. Being far below the indsutry standard of 5% significance level, it can be concluded that at least one of the features is useful in predicting the response variable (ISI). Therefore the null hypothesis $H_{0}$ is rejected.
- The model yields an r-squared value of 0.66, indicating that it would likely be moderately useful in predicting the initial rate of spread of a fire. 

### 4. Cooks distance
After fitting the model, cook's distance  is utilized to analyze if any influential data points affect the regression line. If any exist, this  decreases the model's ability to generalize. 

Based on the plot below, it can be confirmed that the training dataset contains no outlier points (>1), or significantly influential outliers.
``` {r}
#cooks distance
library(stats)
cook = cooks.distance(m)   # here m is the linear mode
plot(cook, ylab = 'cooks dist')
```

### 5. Testing the model

The model can be tested, and resulting mean squared error support our conclusion that the model is moderately useful. 
``` {r}
sum_square_errors <- sum((test$trans - predict(m,newdata = test))^2)
sum_square_errors
```

``` {r}
## Code archives

week = function(x){
  if(x == 'mon'){
    return(1)
  } else if(x == 'tue'){
    return(2)
  } else if(x == 'wed'){
    return(3)
  } else if(x == 'thu'){
    return(4)
  } else if(x == 'fri'){
    return(5)
  } else if(x == 'sat'){
    return(6)
  } else{
    return(7)
  }
}


#check correlation and distributions
#pairs.panels(fire)


#summary(lm(trans ~ month + FFMC + temp + RH + wind, data = fire))
#displaying first 5 rows

```

Footnotes: 

[^1]: http://cwfis.cfs.nrcan.gc.ca/background/summary/fwi <br>
[^2]: http://cwfis.cfs.nrcan.gc.ca/background/summary/fwi <br>
[^3]: https://www.iafss.org/publications/aofst/1/77/view/aofst_1-77.pdf <br>
[^4]: https://cfs.nrcan.gc.ca/publications/download-pdf/28538 <br>
[^5]: https://www.bbc.com/news/world-europe-45070498 <br>






