---
title: "Predicting Fire Behavior Using Weather & Meteorological Data"
author: "Darshil Desai, Edwin Ramirez"
date: "November 27, 2018"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

# 2.1 Abstract

# 2.2 Introduction
Our project will focus on utilizing previous forest fire and weather data sourced from the UCI Machine Learning Repository to predict the Initial Spread Index (ISI) of fires in Portugal and any other countries using the Canadian Forest Wildfire Index System. We found this project to be of relevance considering the recent devastating droughts and fires in Northern Europe throughout Greece, Portugal, and Spain within the last year. Additionally, the relevance of the recent California fires also give support to the necessity of predictive analytics in this area. Hence, by utilizing a regressional model, we can predict the early behavior of a fire. Portugal utilizies the Canadian Forest Fire Weather Index System in order to track fuel moisture and wind to determine the intensity of a fire. Thus, by predicting the initial spread index of potential fires, we'd be able to predict the danger of a fire based on how quickly it would spread. The ISI scale begins at 0, where a 10 indicates a high rate of spread after ignition, and 16 or higher indicates an extreme rapid rate of spread.   

The research question can be hypothesized aa follows: 
H0: None of predictor variables in the dataset are useful in making predictions about the Initial Fire Index
ha: At least one of predictor variables in the dataset is useful in making predictions about the Initial Fire Index


# 2.3 Dataset 

## View Dataset

Lets take a look at the features in our dataset
Our dataset comprises of the following 13 variables: 

1 & 2. X,Y : coordinates within the Montesinho park. Ranges from 1 to 9 <br>
3. month: month when the fire frst occured <br>
4: day: day of the given month when the fire occured <br>
5: FFMC (Fine Fuel Moisture Code) : a numeric rating of the moisture content of litter and other cured fine fuels. This code is an indicator of the relative ease of ignition and the flammability of fine fuel. <br>
6: DMC (Duff Moisture Code): A numeric rating of the average moisture content of loosely compacted organic layers of moderate depth. This code gives an indication of fuel consumption in moderate duff layers and medium-size woody material.[^1] <br>
7.Burned area of the forest: area of forest burned  <br>
8. temperature <br>
9: RH (relative humidity) <br>
10: wind: wind speed <br>
11: rain <br>
12: DC (Drought Code):  A numeric rating of the average moisture content of deep, compact organic layers. This code is a useful indicator of seasonal drought effects on forest fuels and the amount of smoldering in deep duff layers and large logs. <br>
13. ISI (Initial Spread Index): expected rate of fire spread.This variable will be our response variable and we will try and establish a linear relationship between the myriad of weather and meterological factors of the forest experiencing fires and the future expected area burn. [^2] <br> 


``` {r see_data, warning = FALSE, echo = FALSE}
#reading in all the packages reqired for the report
library(ggcorrplot)
library(dplyr)
library(corrplot)
library(leaps)
library(dummies)
library(car)
library("psych")

#reading in the csv file
fire <- read.csv('forestfires.csv') 
head(fire)
```

## 2.4 Methodology
In order to draw a relationship between the various weather, meteorological variables in our dataset and the response variable (Initial Spread Index), it is vital to transform and preprocess our data to succesfully represent this data in a linear relationship. 

Our methods to do the same involve the following: <br>
1. Deleting redundant data that will fail to enhance the model's predictive power <br>
2. Representing categorical data (months) as binary numerical data. <br>
3. Transforming our response variable <br>
4. Examine multicolinearity <br>

### 1. Deleting Redundant Data
Several variables in the given dataset do not assist in creating the linear relationship between with the response variable (Initial Spread Index). Briefly:<br> 
a. X,Y: Removal of coordinate data as this limits the model to the specific area of the forest in the Portugal and does not generalize well to other regions of the world. For instance, if this mode were to be used elsewhere the coordinate data of its forest region would be redundant <br>
b. Days: In isolation, the numerical value of a day does not provide any significant relevance to a datapoint <br>

``` {r delete_data}
fire <- fire[,-match(c("day", 'X', 'Y'),names(fire))]
```


### 2. Convertion categorical data into numerical representations
 We will go ahead and represent the 12 months into 4 seasons as we deem it necessary to our analysis of seasonal drought. Furthermore, the 4 seasons will then be represented as 4 features comprising binary values indicating what season did the fire occur in 
 
```{r data_cleaning_two}
spring <- c('mar', 'apr', 'may')
summer <- c('jun', 'jul', 'aug')
autumn <- c('sep', 'oct','nov')
winter <- c('dec','jan', 'feb')

#clean data
# Representing months as the appropriate season it is a part of
season = function(x){
  if(x %in% spring){
    return('spring')
  } else if(x %in% summer){
    return('summer')
  } else if(x %in% autumn){
    return('autumn')
  } else{
    return('winter')
  }
}

fire$month = sapply(fire$month, season)

#renaming the month column to "seasons"
colnames(fire)[colnames(fire)=="month"] <- "season_"

#Converting the months column into 4 columns, one for each season
fire <- dummy.data.frame(fire)
invisible(get.dummy( fire, 'season_'))

```

### 3.Transforming the response variable
As we can see from the histogram below, the distribution of our data is right-skewed. Therefore, to solve this issue we transform the data, so that a normality violation won't affect the estimates of the standard error. We transform the data by taking the log of all our ISI data.  

```{r show}
#check normality of response variable
hist(fire$ISI, col = "lightblue", xlab = "ISI", main = "Histogram of ISI")
```


```{r transform}
#transform data by taking log
trans = log(fire$ISI)

#change  the below case or error occurs
trans[380] = 1.2

hist(trans, xlab = "Transformed ISI", col = "lightblue", main = "Histogram of ISI Transformed")
abline(v = mean(trans), col = 'red')
abline(v = median(trans), col = 'blue')

```

Consequentially, the distribution the response variable (ISI) is now normal, and we can proceed with selecting the appropriate features for our model. A QQ-normal plot is shown below to illustrate that our data is normally distributed.  

```{r}
#use QQ plot check model
summary(lm(trans ~.-ISI, data = fire))
plot(lm(trans ~.-ISI, data = fire),2)
```

### 4.Examining Multicolinearity {.tabset}
Multicolinearity occurs when predictor variables are linearly correlated with the other. This implies that a change in any one of the predictor variables would entail a change in another highly correlated predictor variable. 

Before we proceed to fit our model it is important to perform 2 vital checks. Firstly, there needs to be a reasonable correlation between the predictor variables and the response variables. An absence of any linear pattern does not warrant the use of a linear regression problem in the first place

#### Colinearity Between Predictor Variables

``` {r}
#Checking for multicolinearity. Here we will compare each of our predictor variables with all the others predictor variables. 
corr<-cor(fire)
#plotting correlation head map
corrplot(corr, type = 'upper')
```
#### Correlation strength Between Predictor Variables and Response Variable 
``` {r analysis, warning = FALSE}
#1. Visualizing correlations between our response variable and the predictor variable 
# Comparing Y with all other predictor variables *(X's)
corr1<-cor(x = fire[,-8], y = trans, use="complete.obs")
ggcorrplot(corr1)
```

The visual analysis above shows us that there are over 8 predictor variables that possess a significant correlation strength with the response variable (transformed ISI). Furthermore we also see that there exists no significant multicolinearity between the predictor variables. In general there always lies the possibility that variables within the same domain will correlate with one another to an extent. 

However to further confirm our claim, we will employ the use of the Variance Inflation Factor analysis. This analysis allows us to support / reject our claim using numerical proof. It is important to note that lower the VIF (lowest being 1), the less multicolinearity exists in our dataset. A VIF of around 5 represents industry standard acceptance rate for multicolinearity as a small value indicates that the standard deviation of the respective variable parameter will remain relatively stable when other predictor variables are added into the regression equation. 

``` {r vif}
#we will go ahead and calculate the Variance Infation Factor for each of our predictor variables
model <- lm(trans~ fire$FFMC + fire$DMC + fire$DC + fire$temp + fire$RH + fire$wind + fire$rain + fire$area, data = fire)
#summary(model)
vif(model)

```
Based on our data above, we can confirm that there exists no significant multicolinearity in our dataset. Most of our VIF values are under 2.5 far below the industry threshold of 5. 

# 2.5 Results & Discussion
Proceeding to establish a sound linear relationship between the weather, meteorological variables and the response variable our next steps comprise the following:  <br>
1. Feature Selection <br>
2. Model fitting using Training Data <br>
3. Analysis <br>

### 1.Feature Selection 
Using stepwise regression, we iterate through different possibilities of a linear model and choose one with features yileding the best (lowest) error metric.

The QQ-normal plot below illustrates the distribution of our data after utilizing stepwise regression for feature selection. 
```{r}
# forward stepwise regression
model = lm(trans ~.-ISI, data = fire)
new_model = step(model)

#new model QQ plot
plot(new_model, 2)

new_model$terms
```


### 2. Fitting the model

Before we go ahead and fit our model, it is important to disproportionally and randomly split up the dataset into training and test data. This will allow us to measure the accuracy of the dataset using the selected variables. 
``` {r}
#attaching the trans column to the main fire dataset
fire_to_split <- cbind(fire, trans)

#Get 400 random numbers
ran <- sample(1:517, 400, replace=F)

#splitting the dataset into training and testing
train <- fire_to_split[ran,]

#test data: contains data not in training data
test = setdiff(fire_to_split, train)
```


Now we will fit a model using the training data with the selected features. After removing features not selected from stepwise regression, our model include `season_autumn`,`season_spring` ,`season_summer`,`FFMC`, `temp`, `RH`, and `wind`
``` {r}
#fitting the model
m = lm(train$trans ~ season_autumn + season_spring + season_summer + FFMC + temp + RH + wind - ISI - trans, data = train)
summary(m)
```

### 3. Analysis
We can make the following conclusions: 
- The overall P-value of the model is <2.2e-16. Being far below the indsutry standard 5% alpha rate we can conclude that at least one of the features is useful in predicting the response variable (ISI)
- The model yields an r-squared value of 0.66, indicating that it would likely be moderately useful in predicting the initial rate of spread of a fire.   

### 4. Cooks distance
After fitting the model, it may be interesting to also look at the cooks distance metrics for our sample datapoints and examine if the regression line is affected, if at all, by amny influential datapoints thus leading decreasing the model's power to generalize. 

Based on the plot below, we can confirm that our training dataset comprises no outlier points (>1) that also happen to be significantly influential outliers.
``` {r}
#cooks distance
library(stats)
cook = cooks.distance(m)   # here m is the linear mode
plot(cook, ylab = 'cooks dist')
```

### 5. Testing the model
we will go ahead and test the model and return the mean squared error of the same
``` {r}
sum_square_errors <- sum((test$trans - predict(m,newdata = test))^2)
sum_square_errors
```

``` {r}
## Code archives

week = function(x){
  if(x == 'mon'){
    return(1)
  } else if(x == 'tue'){
    return(2)
  } else if(x == 'wed'){
    return(3)
  } else if(x == 'thu'){
    return(4)
  } else if(x == 'fri'){
    return(5)
  } else if(x == 'sat'){
    return(6)
  } else{
    return(7)
  }
}


#check correlation and distributions
#pairs.panels(fire)


#summary(lm(trans ~ month + FFMC + temp + RH + wind, data = fire))
#displaying first 5 rows

```

Footnotes: 

[^1]: http://cwfis.cfs.nrcan.gc.ca/background/summary/fwi <br>
[^2]: http://cwfis.cfs.nrcan.gc.ca/background/summary/fwi <br>
[^3]: https://www.iafss.org/publications/aofst/1/77/view/aofst_1-77.pdf <br>
[^4]: https://cfs.nrcan.gc.ca/publications/download-pdf/28538 <br>
[^5]: https://www.bbc.com/news/world-europe-45070498 <br>






